
---
title: "VarClust package tutorial"
author: "Piotr Sobczyk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VarClust tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

VarClust is a package that enables dimension
reduction via variables clustering. Classical methods for
dimensionality reduction are based on removing variables that
are redundant. Here we assume that for certain group of variables 
one can define few latent variables that describe our data.

This tutorial will gently introduce you to usage of package **VarClust** and
familiarize with its options.

You can install VarClust from either CRAN (stable version) or github (the current development version).

```{r, results='hide'}
library(VarClustPoprawka)
library(mclust)
```

First let us consider some real genomic data. We're going to use FactoMineR package data.
This data consists of two types of variables. First group are gene expression data.
The second is RNA data.
Please note that it may take few minutes to run the following code:
```{r}
comp <- read.table("http://factominer.free.fr/docs/gene.csv",sep=";",header=T,row.names=1) 
benchmarkClustering <- c(rep(1, 68), rep(2, 356))    
comp <- data.frame(comp[,-ncol(comp)])    
mlcc.fit <- mlcc.bic(comp, numb.clusters = 1:3, numb.runs = 20, max.dim = 3, greedy=F, estimate.dimensions=T, numb.cores=1)
summary(mlcc.fit)
mclust::adjustedRandIndex(mlcc.fit$segmentation, benchmarkClustering)
```

If you know what is the true number of clusters you might use **mlcc.reps** function instead.
```{r}
mlcc.fit2 <- mlcc.reps(X=comp, numb.clusters = 2, numb.runs = 20, max.dim = 4, numb.cores=1)
mclust::adjustedRandIndex(mlcc.fit2$segmentation, benchmarkClustering)
```
You should also use **mlcc.reps** function if you have some apriori knowledge regarding true segmentation.
You can enforce starting point
```{r}
mlcc.fit3 <- mlcc.reps(comp, numb.clusters = 2, numb.runs = 0, max.dim = 2, initial.segmentations=list(benchmarkClustering), numb.cores=1)
mclust::adjustedRandIndex(mlcc.fit3$segmentation, benchmarkClustering)
```
